---
title: "Recommender System Trial: Books"
author: "Joseph Lisle"
output: 
  html_document:
    theme: Superhero
    highlight: zenburn
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
---

# Introduction


# Setting up work space
## Loading Packages
```{r}
library(dplyr)
library(tidyr)
library(shiny)
library(shiny.semantic)
```

## Loading data into work space
```{R}
Book.Tags <- read.csv("book_tags.csv")
Books <- read.csv("books.csv")
Ratings <- read.csv("ratings.csv")
Tags <- read.csv("tags.csv")
To.Read <- read.csv("to_read.csv")
```

## Preliminary look at data
```{r}
head(Book.Tags)
head(Books)
head(Ratings)
head(Tags)
head(To.Read)
```

# Genres
## Counts of Tags
```{r}
# Creating a data frame that has the count per book tag
Book.Tags.Count <- Book.Tags %>%
  group_by(tag_id) %>%
  summarize(Count = sum(count))

# Merging in the tag names
Book.Tags.Count <- merge(x=Book.Tags.Count, y=Tags, by="tag_id", all=TRUE)

# Arranging the data frame based on number of tags
Book.Tags.Count <- Book.Tags.Count %>%
  arrange(desc(Count))

# Exporting data so I can specify in Excel which tags I want to keep
write.csv(Book.Tags.Count, "Book Tags count.csv", row.names=FALSE)
```

## Importing Fixed Tags
```{r}
# I went into Excel and looked at every tag that was used more than 10k times.
# After limiting this list, I then grouped the tags that were similar but did not have the same spellings (i.e. "young adult" and "ya" were grouped together)

# Uploading the fixed tags
Book.Tags.Fixed <- read.csv("Book Tags Fixed.csv")
```

## Limiting Top Tags for Each Book
```{r}
# Now that I have the top tags in general, I am going to remove any tags from the Book Tags dataset that is not within the top tags list that I created
# I'll perform an inner join on the Book.Tags and Book.Tags.Fixed dataset to remove any tags that aren't within both datasets
Popular.Book.Tags <- merge(x=Book.Tags, y=Book.Tags.Fixed, by="tag_id")

# Cleaning up Popular.Book.Tags so that it only contains the necessary information
Popular.Book.Tags <- Popular.Book.Tags %>%
  select(goodreads_book_id, genre = Altered.Tag.Name, count) %>% #selecting necessary variables
  group_by(goodreads_book_id, genre) %>% #grouping by book id and genre to collect counts together for genres with same name
  summarize(count = sum(count)) %>%
  arrange(goodreads_book_id, desc(count)) %>% #removing any genre that isn't within top 3 per book
  top_n(n=3, wt=count)

# Adding in a rank variable so that I can specify which genres have more counts
Popular.Book.Tags <- Popular.Book.Tags %>%
  group_by(goodreads_book_id) %>%
  mutate(rank = row_number(-count))

# confirming everything looks as expected
head(Popular.Book.Tags, 30)
```

## Assigning top tags by book
```{r}
# Creating a dataset that specifies the top tags by book
Popular.Tags.by.Book <- Popular.Book.Tags %>%
  select(-count) %>%
  spread(rank, genre) %>%
  select(goodreads_book_id, "Genre 1" = "1", "Genre 2" = "2", "Genre 3" = "3")

# Confirming things look as expected
head(Popular.Tags.by.Book, 30)
```

## Adding Genres to Books Data Frame and Adding Buckets for Number of Ratings
```{r}
# Now that I have the popular genres by book, I will add it to that data frame to use in other analysis/recommending
Books <- merge(x=Books, y=Popular.Tags.by.Book, by="goodreads_book_id")

# Assigning bucket categories for number of ratings
Books$Rating.Bucket <- ifelse(Books$ratings_count > 100000, "100k+",
                       ifelse(Books$ratings_count > 50000, "50k - 100k",
                       ifelse(Books$ratings_count > 25000, "25k - 50k",
                       ifelse(Books$ratings_count > 15000, "15k - 25k", 
                       ifelse(Books$ratings_count > 10000, "10k - 15k", "<10k")))))

# Creating a list of the buckets for drop down menu
Rating.Buckets <- Books %>%
  select(Rating.Bucket) %>%
  group_by(Rating.Bucket) %>%
  summarize(Count = n()) %>%
  select(Rating.Bucket) %>%
  as.list()

# Assigning bucket categories for year of publication
Books$Publication.Year.Bucket <- ifelse(Books$original_publication_year < 1800, "Before 1800",
                 ifelse(Books$original_publication_year < 1900, "1800 - 1900",
                 ifelse(Books$original_publication_year < 1925, "1900 - 1925", 
                 ifelse(Books$original_publication_year < 1950, "1925 - 1950",
                 ifelse(Books$original_publication_year < 1970, "1950 - 1970",
                 ifelse(Books$original_publication_year < 1980, "1970s",
                 ifelse(Books$original_publication_year < 1990, "1980s",
                 ifelse(Books$original_publication_year < 2000, "1990s",
                 ifelse(Books$original_publication_year < 2005, "2000 - 2005",
                 ifelse(Books$original_publication_year < 2010, "2005 - 2010", "2010 - 2017"))))))))))

# Creating a list of publication buckets for drop down menu in app
Publication.Year.Buckets <- Books %>%
  select(Publication.Year.Bucket) %>%
  group_by(Publication.Year.Bucket) %>%
  summarize(Count = n()) %>%
  select(Publication.Year.Bucket) %>%
  as.list()

# Confirming that it looks as expected
summary(Books)
```

## Attaching Book Genres to Ratings
```{r}
# Creating data frame with book information attached to ratings by user
Ratings.with.Genres <- merge(x=Ratings, y=Books, by="book_id")

# Keeping only the necessary information and preparing to create multiple 
Ratings.with.Genres <- Ratings.with.Genres %>%
  select(user_id, book_id, rating, "Genre 1", "Genre 2", "Genre 3")

# Confirming everything looks good to go
head(Ratings.with.Genres)

# Converting dataset from wide to long (haha it's already almost 6Mil rows long, but even longer is what I mean)
Ratings.with.Genres <- Ratings.with.Genres %>%
  gather("Genre Number", Genre, -user_id, -book_id, -rating)

# Confirming things look good to go
head(Ratings.with.Genres)
```

## Compiling Favorite Genres by User
```{r}
# Creating a dataset that has top genres per user by count of books read per genre
Top.User.Ratings.by.Genre.Count <- Ratings.with.Genres %>%
  group_by(user_id, Genre) %>%
  summarize(count = n(), Mean_Rating = mean(rating)) %>%
  arrange(user_id, desc(count)) %>%
  top_n(n=3, wt=count) %>%
  mutate(rank = row_number(-count))

# Creating a dataset that has top genres per user by rating of books in that genre (if they've read 10+ books in that genre)
Top.User.Ratings.by.Rating.Rank <- Ratings.with.Genres %>%
  group_by(user_id, Genre) %>%
  summarize(count = n(), Mean_Rating = mean(rating)) %>%
  filter(count > 10) %>%
  arrange(user_id, desc(Mean_Rating)) %>%
  top_n(n=3, wt=Mean_Rating) %>%
  mutate(rank = row_number(-Mean_Rating))

# Checking to ensure the data looks as expected
head(Top.User.Ratings.by.Genre.Count)
head(Top.User.Ratings.by.Rating.Rank)

# Showing the genres
table(Top.User.Ratings.by.Genre.Count$Genre) #64 genres
table(Top.User.Ratings.by.Rating.Rank$Genre) #95 genres

# Creating a dataset from which I can pull the genres in order of count of occurences
Top.User.Ratings.by.Rating.Rank.Vec <- Top.User.Ratings.by.Rating.Rank %>%
  group_by(Genre) %>%
  summarize(Count = sum(count)) %>%
  select(Genre, Count) %>%
  arrange(desc(Count))

# Assigning the vector to just the genres, so that I can input this into my selection function
Top.User.Ratings.by.Rating.Rank.Vec <- as.vector(Top.User.Ratings.by.Rating.Rank.Vec$Genre)
Top.User.Ratings.by.Rating.Rank.Vec

# Creating a dataset from which I can pull the genres in order of number of users with that genre in their top 3 favorites. Each book should be averaged from at least 11 users, so I am removing any genres with fewer than 11 users with that genre in their top 3.
Top.User.Ratings.by.Rating.Rank.Vec.Short <- Top.User.Ratings.by.Rating.Rank %>%
  group_by(Genre) %>%
  summarize(Count = n()) %>%
  arrange(desc(Count)) %>%
  filter(Count>10)

# Assigning a shortened vector of genres to allow for more robust reporting. 
Top.User.Ratings.by.Rating.Rank.Vec.Short <- as.vector(Top.User.Ratings.by.Rating.Rank.Vec.Short$Genre)
```

## Attaching Favorite User Genres to Book Reviews
```{r}
# Creating data frame that has favorite genres for each user review
Ratings.with.Favorite.User.Genres.Rating.Rank <- merge(x=Ratings, y=(Top.User.Ratings.by.Rating.Rank %>% select(user_id, Genre)), by="user_id")

# Creating a list of mean ratings based on users' favorite genres (grouped at the book and genre level)
Book.Ratings.with.Favorite.User.Genres.Rating.Rank <- Ratings.with.Favorite.User.Genres.Rating.Rank %>%
  group_by(book_id, Genre) %>%
  summarize(Mean_Rating = mean(rating), count=n()) %>%
  filter(count>10)

# Checking to make sure it's working correctly
head(Book.Ratings.with.Favorite.User.Genres.Rating.Rank)
```

## Creating CSVs of Data for Easy Pickup
```{r}
# While working on the project, I got tired of running my code every time I started working again, so I wrote CSVs for all the files I would need.
write.csv(Top.User.Ratings.by.Genre.Count, file="Top User Ratings by Genre Count.csv", row.names = FALSE)
write.csv(Top.User.Ratings.by.Rating.Rank, file="Top User Ratings by Rating Rank.csv", row.names = FALSE)
write.csv(Ratings.with.Genres, file="Ratings with Genres.csv", row.names=FALSE)
write.csv(Books, file="Books with Genres.csv", row.names=FALSE)
write.csv(Top.User.Ratings.by.Rating.Rank.Vec.Short, file="List of Popular Genres based on Rating Rank.csv", row.names=FALSE)
write.csv(Book.Ratings.with.Favorite.User.Genres.Rating.Rank, file="Book Ratings with Favorite User Genres by Rating Rank.csv", row.names=FALSE)
```

## Loading CSVs for Easy Pickup
```{r}
# Just reading in the CSV files that I saved so I can pick up again easily.
Top.User.Ratings.by.Genre.Count <- read.csv("Top User Ratings by Genre Count.csv")
Top.User.Ratings.by.Rating.Rank <- read.csv("Top User Ratings by Rating Rank.csv")
Ratings.with.Genres <- read.csv("Ratings with Genres.csv")
Books <- read.csv("Books with Genres.csv")
Ratings <- read.csv("Ratings.csv")
Top.User.Ratings.by.Rating.Rank.Vec.Short <- read.csv("List of Popular Genres based on Rating Rank.csv")
Book.Ratings.with.Favorite.User.Genres.Rating.Rank <- read.csv("Book Ratings with Favorite User Genres by Rating Rank.csv")

# Creating list of publication year buckets to use in app
Publication.Year.Buckets <- Books %>%
  select(Publication.Year.Bucket) %>%
  group_by(Publication.Year.Bucket) %>%
  summarize(Count = n()) %>%
  select(Publication.Year.Bucket) %>%
  as.list()

# Creating list of number of ratings buckets to use in app
Rating.Buckets <- Books %>%
  select(Rating.Bucket) %>%
  group_by(Rating.Bucket) %>%
  summarize(Count = n()) %>%
  select(Rating.Bucket) %>%
  as.list()

# Creating a list of top user ratings in order of most often assigned
Top.User.Ratings.by.Rating.Rank.Vec.Short <- Top.User.Ratings.by.Rating.Rank.Vec.Short %>%
  arrange(desc(Count)) %>%
  select(Genre) %>%
  as.list()
```

